Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-05 MScDiplom_Lemenkova.tex
comment 10 of Valentijn: added legend to GNUplot-18.

	\vspace{1ex}
	\textbf{Main Tools}:
	\begin{itemize}
		\item \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS 10.0}: mapping, data integration and spatial analysis
		\item \href{http://www.ilwis.org/}{ILWIS}, \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine}: imagery processing and classification
		\item \href{http://www.filewatcher.com/b/ftp/ftp.dfd.dlr.de/pub/WASI.0.0.html}{WASI Water Color Simulator}, for modelling optical water 					properties under changing environment
		\item \href{http://welcome.hp.com/country/us/en/prodserv/handheld.html}{iPAQ} and \ac{GPS} for data capture during the fieldwork
		\item \href{http://www.bibtex.org/de/}{BibTeX}, for management of bibliography and reference database
		\item \href{http://www.latex-project.org/}{\LaTeX}, for document typesetting, formatting and performing pdf output
		\item \href{http://www.python.org/}{Python language}, for writing auxiliary scripts (e.g. raw data interpolation)
		\item \href{http://www.gnuplot.info/}{Gnuplot} for plotting graphs
		\item \href{http://www.openoffice.org/}{Open Office}: document typesetting, data pre-processing and preliminary statistical evaluation

		\item \href{http://www.spss.com/}{SPSS (PASW Statistics 18)}: statistical data edition and analysis 
		\item \href{http://gretl.sourceforge.net/}{Gretl: Gnu Regression, Econometrics and Time-series Library} for statistic analysis
		\item \href{http://staruml.sourceforge.net/en/}{StarUML}, \href{http://live.gnome.org/Dia}{Dia}, \href{http://argouml.tigris.org/}						{ArgoUML}: \ac{UML} diagram tool for creating flowcharts of work algorithms
		\item \href{http://inkscape.org/}{Inkscape}: drawing
	\end{itemize}

	\vspace{1ex}
	\textbf{Auxiliary Tools}:
	\begin{itemize}
		\item \href{http://www.lyx.org/}{LyX} for exporting \LaTeX document to the Word .doc format via HTML
		\item \href{http://calc2latex.sourceforge.net/}{Calc2LaTeX}: converting spreadsheets into \LaTeX environment
		\item \href{http://openev.sourceforge.net/}{OpenEV}: used for raster Google Earth imagery viewing, visualisation and preliminary analysis
		\item \href{http://www.notepad-plus-plus.org/}{Notepad++}: auxiliary text editing and raw data formatting
	\end{itemize}

@@

\subsection{Research approach}\index[concepts]{Research!approach}
Seagrass consistent monitoring and mapping is necessary and important for the sustainable coastal
development and conservation measures. Earlier, many seagrass meadows have been destroyed by
human activities in the coastal zone, mainly due to the ignorance of their existence, because
information on the seagrass bed exact location was not available \cite{Choo06}\label{Choo06}. Well-time seagrass
observations and mapping enables precise control of its spatial distribution, detection of any changes
in the seagrass landscapes, highlights potential environmental threats in the coastal zone (e.g.
declining of meadows) before they become unmanageable for the coastal management services.
Choosing the right and most effective approach method for the seagrass monitoring is essential.
Remote sensing methods alone, though having evident advantages, are insufficient, because satellite
images of underwater habitats are notoriously difficult to identify and interpret. The best research
method should be based on the integrated approach, well described in various scientific works \cite{Brown02,Montefalcone06,Kirkman96}\label{Brown02} \label{Montefalcone06} \label{Kirkman96}, which includes combination of various
techniques of the seagrass monitoring, i.e. remote sensing imagery classification of aerial and satellite
images, GIS-based spatial analysis and ground \textit{in-situ} surveys.
\begin{figure}[H]
	\centering
	\includegraphics[scale=0.48]{UML_ClassDiagram.jpg}
	\caption{General methodological approach for the analysis of spectral signatures. \href{http://argouml.tigris.org/}{ArgoUML}}
	\label{fig:1.10}
\end{figure}
The current study is based on the application of the remote sensing data, broadband satellite imagery,
aerial images and the results of the underwater videographic measurements towards seagrass
mapping  (Fig.\ref{fig:1.11}). Image classification is based on the principle of the differentiation between the spectral
signatures of various seafloor cover types  (Fig.\ref{fig:1.10}). The spectrum of light coming up from the ocean surface in
shallow waters keeps information on the optical properties of the seawater components and benthic
substrate which can be read from their spectral signatures \cite{Werdell03}\label{Werdell03}. \\
The spectral reflectance, distinct for each seafloor cover type, enables their discrimination on the image. The pre-processing
of the images includes imagery corrections for atmospheric noises and effects of the water
column. Reflectance spectra of the seagrass canopy at different depths of the water-column are
analysed for the discrimination of their spectral signatures, enabling to separate various seafloor types
during classification. The results of the of imagery classification are analysed for the detection of the
dynamics in \textit{P.oceanica} seagrass distribution along the northern coasts of Crete. Aerial imagery from
Google Earth with high spatial resolution, suitable for the large-scale detailed mapping of seagrass
mattes, is used for the improvement of the accuracy of large seagrass meadows and separate mattes
within the meadows. The \textit{in situ} underwater videometric measurements of the seafloor are collected
during the fieldwork in Crete, for the validation of the classification results and to determine the exact
current distribution of the \textit{P.oceanica} meadows. The image processing includes steps of the remote
sensing techniques, i.e. calibration, masking from land and cloud, atmospheric correction, sea surface
glint and depth effects correction as recommended \cite{Matarrese08}\label{Matarrese08}. During the image classification
working step the training sites for the supervised classification methods are designed, as well as its
control and trials of different classification approaches (Unsupervised, K-means or Isodata;
Supervised, Maximum Likelihood).

@@

@@

The backward-scattering coefficient (bb), also included in WASI, is a
fundamental optical property which plays a central role in the ocean-color remote sensing, providing
the remotely sensed optical signal, as well as suspended particle distributions, water clarity, and
underwater visibility \cite{Maffione97}\label{Maffione97}. WASI enables simulation of backscattering of pure
water, large and small particles. The values of all these parameters can be redacted and changed
manually. However, the most important, major factors affecting the \textit{in-situ} conditions are water depth (Fig.\ref{fig:4.9})
and cover fraction of the seafloor types: \textit{P.oceanica} and carbonate sand.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.27]{GNU-18.jpg}
	\caption{Multiplot showing spectral reflectance of the seawater with sediments, measured in aquarium tank, Agia Pelagia district, Crete. \href{http://www.gnuplot.info/}{Gnuplot} \\Two complimentary graphs below show the results of the statistical analysis.}
	\label{fig:4.4}
\end{figure}

Seagrass \textit{P.oceanica} can be mapped using remotely acquired spectral imagery, but only if it has distinctive
reflectance signatures at different depths. Therefore, the depths of the shelf area are the first variable
condition for this research question. The depths values chosen for the current research are lesser than
3.5 meters, covering shelf zone, and providing the best environmental conditions for the seagrass
\textit{P.oceanica}: 0.5, 2.0 and 3.5 meters with an interval of 1.5 meters.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.3]{GNU-20-new.jpg}
	\caption{Statistical comparison of spectral reflectance of \textit{P.oceanica} and carbonate sand. Selected measurements sets. \href{http://www.gnuplot.info/}{Gnuplot} graph}
	\label{fig:28}
\end{figure}

The 3.5 m depth limit was chosen based on the analysis of the separability of seagrass reflectance signatures, received by the means of
previous \textit{in-situ} measurements (year 2009) of the radiance and irradiance of water in Agia Pelagia
bay, which indicated that \textit{P.oceanica} seagrass could be only well separated within depths of 3.5 m.
A statistical analysis of WASI-simulated spectral reflectance has been used in order to answer the first
research question: whether the 	extit{P.oceanica} spectra is spectrally distinct at varying environmental \textit{in-situ}
conditions, and if \textit{P.oceanica} remains spectrally distinct with the increasing water column depth.
To answer this research question, different seafloor cover types are discriminated using data of the
broadband remote sensing.

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.40]{GNU-19.jpg}
	\caption{Statistical analysis of the spectral reflectance of \textit{P.oceanica}: min-max, average values (red bold points), Q1-Q3 areas (green vertical dashed) and 			measured values (dotted): multiplot of measurement sets 200-300. \href{http://www.gnuplot.info/}{Gnuplot}}
	\label{fig:4.10}
\end{figure}
\begin{figure}[H]
	\centering
	\subfloat [Sand]{\label{fig:34a}\includegraphics[width=0.4\textwidth]{Fig-27a.jpg}}
	\subfloat [\textit{	extit{P.oceanica}}]{\label{fig:34b}\includegraphics[width=0.4\textwidth]{Fig-27b.jpg}}
	\caption{Bottom albedo of carbonate sand and \textit{P.oceanica}, Agia Pelagia}
	\label{fig:4.11}
\end{figure}

@@

\begin{table}
\rowcolors{1}{Khaki}{LightYellow}
\caption{Statistical analysis of the measurements of spectral reflectance of 	extit{P.oceanica} (fragment)}
\centering
  \begin{tabular}{| c | c | c | c | c |}
    \hline
   	 \textbf{Statistical values} & \textbf{318.19} & \textbf{418.3448} & \textbf{518.8361} & \textbf{619.3662} \\ \hline \hline
	    Mean & 0.034521 & 0.049235 & 0.076088 & 0.085287 \\ \hline
	    St Dev & 0.019849 & 0.030338 & 0.042587 & 0.046569 \\ \hline
	   Median & 0.033035 & 0.046197 & 0.07704 & 0.080277 \\ \hline
	   Q1 & 0.019293 & 0.023706 & 0.045606& 0.049468 \\ \hline
	   Q3 & 0.042477 & 0.057934 & 0.092289 & 0.114191 \\ \hline
	   Minimum & 0.007898 & 0.010087 & 0.021373 & 0.022691 \\ \hline
	   Maximum & 0.076835 & 0.110906 & 0.18938 & 0.216231 \\ \hline
	   25th Pct & 0.019293 & 0.023706 & 0.045606 & 0.049468 \\ \hline
	   50th Pct & 0.013743 & 0.022491 & 0.031435 & 0.030808 \\ \hline
	   75th Pct & 0.009441 & -0.01174 & -0.01525 & -0.03391 \\ \hline
	   Min & 0.011395 & 0.013618 & 0.024232 & 0.026777 \\ \hline
	   Max & 0.034358 & 0.052972 & 0.097091 & 0.1 \\ \hline
  \end{tabular}
   \label{tab:4}
\end{table}

>>>>

	\vspace{1ex}
	\textbf{Main Tools}:
	\begin{itemize}
		\item \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS 10.0}: mapping, data integration and spatial analysis
		\item \href{http://www.ilwis.org/}{ILWIS}, \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine}: imagery processing and classification
		\item \href{http://www.filewatcher.com/b/ftp/ftp.dfd.dlr.de/pub/WASI.0.0.html}{WASI Water Color Simulator}, for modelling optical water 					properties under changing environment
		\item \href{http://welcome.hp.com/country/us/en/prodserv/handheld.html}{iPAQ} and \ac{GPS} for data capture during the fieldwork
		\item \href{http://www.latex-project.org/}{\LaTeX}, for document typesetting, formatting and performing pdf output
		\item \href{http://www.bibtex.org/de/}{BibTeX}, for management of bibliography and reference database
		\item \href{http://www.gnuplot.info/}{GNUplot} for plotting graphs
		\item \href{http://gretl.sourceforge.net/}{Gretl: GNU Regression, Econometrics and Time-series Library} for statistic analysis
		\item \href{http://www.openoffice.org/}{Open Office}: document typesetting, data pre-processing and preliminary statistical evaluation		
		\item \href{http://www.spss.com/}{SPSS (PASW Statistics 18)}: statistical data edition and analysis

		\item \href{http://staruml.sourceforge.net/en/}{StarUML}, \href{http://live.gnome.org/Dia}{Dia}, \href{http://argouml.tigris.org/}						{ArgoUML}: \ac{UML} diagram tool for creating flowcharts of work algorithms
	\end{itemize}

	\vspace{1ex}
	\textbf{Auxiliary Tools}:
	\begin{itemize}
		\item \href{http://www.lyx.org/}{LyX} for exporting \LaTeX document to the Word .doc format via HTML
		\item \href{http://calc2latex.sourceforge.net/}{Calc2LaTeX}: converting spreadsheets into \LaTeX environment
		\item \href{http://www.python.org/}{Python language}, for writing auxiliary scripts (e.g. raw data interpolation)
		\item \href{http://openev.sourceforge.net/}{OpenEV}: used for raster Google Earth imagery viewing, visualisation and preliminary analysis
		\item \href{http://www.notepad-plus-plus.org/}{Notepad++}: auxiliary text editing and raw data formatting
		\item \href{http://inkscape.org/}{Inkscape} and \href{http://www.gimp.org/}{Gimp: GNU Image Manipulation Program} for auxiliary drawing
	\end{itemize}

@@

\subsection{Research approach}\index[concepts]{Research!approach}
Seagrass consistent monitoring and mapping is necessary and important for the sustainable coastal
development and conservation measures. Earlier, many seagrass meadows have been destroyed by
human activities in the coastal zone, mainly due to the ignorance of their existence, because
information on the seagrass bed exact location was not available \cite{Choo06}\label{Choo06}. Well-time seagrass
observations and mapping enables precise control of its spatial distribution, detection of any changes
in the seagrass landscapes, highlights potential environmental threats in the coastal zone (e.g.
declining of meadows) before they become unmanageable for the coastal management services.
Choosing the right and most effective approach method for the seagrass monitoring is essential.
Remote sensing methods alone, though having evident advantages, are insufficient, because satellite
images of underwater habitats are notoriously difficult to identify and interpret. The best research
method should be based on the integrated approach, well described in various scientific works \cite{Brown02,Montefalcone06,Kirkman96}\label{Brown02} \label{Montefalcone06} \label{Kirkman96}, which includes combination of various
techniques of the seagrass monitoring, i.e. remote sensing imagery classification of aerial and satellite
images, GIS-based spatial analysis and ground \textit{in-situ} surveys.
\begin{figure}[H]
	\centering
	\includegraphics[scale=0.48]{UML_ClassDiagram.jpg}
	\caption{General methodological approach for the analysis of spectral signatures. \href{http://argouml.tigris.org/}{ArgoUML}, \href{http://www.gimp.org/}{Gimp}}
	\label{fig:1.10}
\end{figure}
The current study is based on the application of the remote sensing data, broadband satellite imagery,
aerial images and the results of the underwater videographic measurements towards seagrass
mapping  (Fig.\ref{fig:1.11}). Image classification is based on the principle of the differentiation between the spectral
signatures of various seafloor cover types  (Fig.\ref{fig:1.10}). The spectrum of light coming up from the ocean surface in
shallow waters keeps information on the optical properties of the seawater components and benthic
substrate which can be read from their spectral signatures \cite{Werdell03}\label{Werdell03}. \\
The spectral reflectance, distinct for each seafloor cover type, enables their discrimination on the image. The pre-processing
of the images includes imagery corrections for atmospheric noises and effects of the water
column. Reflectance spectra of the seagrass canopy at different depths of the water-column are
analysed for the discrimination of their spectral signatures, enabling to separate various seafloor types
during classification. The results of the of imagery classification are analysed for the detection of the
dynamics in \textit{P.oceanica} seagrass distribution along the northern coasts of Crete. Aerial imagery from
Google Earth with high spatial resolution, suitable for the large-scale detailed mapping of seagrass
mattes, is used for the improvement of the accuracy of large seagrass meadows and separate mattes
within the meadows. The \textit{in situ} underwater videometric measurements of the seafloor are collected
during the fieldwork in Crete, for the validation of the classification results and to determine the exact
current distribution of the \textit{P.oceanica} meadows. The image processing includes steps of the remote
sensing techniques, i.e. calibration, masking from land and cloud, atmospheric correction, sea surface
glint and depth effects correction as recommended \cite{Matarrese08}\label{Matarrese08}. During the image classification
working step the training sites for the supervised classification methods are designed, as well as its
control and trials of different classification approaches (Unsupervised, K-means or Isodata;
Supervised, Maximum Likelihood).

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.35]{GNU-20-new.jpg}
	\caption{Statistical comparison of spectral reflectance of \textit{P.oceanica} and carbonate sand. Selected measurements sets. \href{http://www.gnuplot.info/}{Gnuplot} graph}
	\label{fig:28}
\end{figure}

@@

The backward-scattering coefficient (bb), also included in WASI, is a
fundamental optical property which plays a central role in the ocean-color remote sensing, providing
the remotely sensed optical signal, as well as suspended particle distributions, water clarity, and
underwater visibility \cite{Maffione97}\label{Maffione97}. WASI enables simulation of backscattering of pure
water, large and small particles. The values of all these parameters can be redacted and changed
manually. However, the most important, major factors affecting the \textit{in-situ} conditions are water depth (Fig.\ref{fig:4.9})
and cover fraction of the seafloor types: \textit{P.oceanica} and carbonate sand.
Seagrass \textit{P.oceanica} can be mapped using remotely acquired spectral imagery, but only if it has distinctive
reflectance signatures at different depths. Therefore, the depths of the shelf area are the first variable
condition for this research question. The depths values chosen for the current research are lesser than
3.5 meters, covering shelf zone, and providing the best environmental conditions for the seagrass
\textit{P.oceanica}: 0.5, 2.0 and 3.5 meters with an interval of 1.5 meters. 
The 3.5 m depth limit was chosen based on the analysis of the separability of seagrass reflectance signatures, received by the means of
previous \textit{in-situ} measurements (year 2009) of the radiance and irradiance of water in Agia Pelagia
bay, which indicated that \textit{P.oceanica} seagrass could be only well separated within depths of 3.5 m.
A statistical analysis of WASI-simulated spectral reflectance has been used in order to answer the first
research question: whether the 	extit{P.oceanica} spectra is spectrally distinct at varying environmental \textit{in-situ}
conditions, and if \textit{P.oceanica} remains spectrally distinct with the increasing water column depth.
To answer this research question, different seafloor cover types are discriminated using data of the
broadband remote sensing.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.3]{GNU-18-legend.png}
	\caption{Multiplot showing spectral reflectance of the seawater with sediments, measured in aquarium tank, Agia Pelagia district, Crete. \href{http://www.gnuplot.info/}{Gnuplot} \\Two complimentary graphs below show the results of the statistical analysis.}
	\label{fig:4.4}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.40]{GNU-19.jpg}
	\caption{Statistical analysis of the spectral reflectance of \textit{P.oceanica}: min-max, average values (red bold points), Q1-Q3 areas (green vertical dashed) and 	measured values (dotted): multiplot of measurement sets 200-300. \href{http://www.gnuplot.info/}{Gnuplot}}
	\label{fig:4.10}
\end{figure}
\begin{figure}[H]
	\centering
	\subfloat [Sand]{\label{fig:34a}\includegraphics[width=0.4\textwidth]{Fig-27a.jpg}}
	\subfloat [\textit{P.oceanica}]{\label{fig:34b}\includegraphics[width=0.4\textwidth]{Fig-27b.jpg}}
	\caption{Bottom albedo of carbonate sand and \textit{P.oceanica}, Agia Pelagia}
	\label{fig:4.11}
\end{figure}

@@

\begin{table}
\rowcolors{1}{Khaki}{LightYellow}
\caption{Statistical analysis of the measurements of spectral reflectance of \textit{P.oceanica} (fragment)}
\centering
  \begin{tabular}{| c | c | c | c | c |}
    \hline
   	 \textbf{Statistical values} & \textbf{318.19} & \textbf{418.3448} & \textbf{518.8361} & \textbf{619.3662} \\ \hline \hline
	    Mean & 0.034521 & 0.049235 & 0.076088 & 0.085287 \\ \hline
	    St Dev & 0.019849 & 0.030338 & 0.042587 & 0.046569 \\ \hline
	   Median & 0.033035 & 0.046197 & 0.07704 & 0.080277 \\ \hline
	   Q1 & 0.019293 & 0.023706 & 0.045606& 0.049468 \\ \hline
	   Q3 & 0.042477 & 0.057934 & 0.092289 & 0.114191 \\ \hline
	   Minimum & 0.007898 & 0.010087 & 0.021373 & 0.022691 \\ \hline
	   Maximum & 0.076835 & 0.110906 & 0.18938 & 0.216231 \\ \hline
	   25th Pct & 0.019293 & 0.023706 & 0.045606 & 0.049468 \\ \hline
	   50th Pct & 0.013743 & 0.022491 & 0.031435 & 0.030808 \\ \hline
	   75th Pct & 0.009441 & -0.01174 & -0.01525 & -0.03391 \\ \hline
	   Min & 0.011395 & 0.013618 & 0.024232 & 0.026777 \\ \hline
	   Max & 0.034358 & 0.052972 & 0.097091 & 0.1 \\ \hline
  \end{tabular}
   \label{tab:4}
\end{table}
